% introduction.tex

% Section Title
\section{INTRODUCTION} \label{sec:introduction}

    Intrusion Detection Systems (IDS) are a critical component of modern network security, tasked with distinguishing legitimate traffic from malicious activity such as denial-of-service attacks, network probing, and unauthorized access attempts. 
    A central challenge in this domain is that labeled anomaly data is often scarce or unavailable, motivating the use of anomaly detection methods that learn normality from benign traffic alone.

    This laboratory investigates supervised, semi-supervised, and unsupervised anomaly detection techniques applied to a network traffic dataset derived from the KDD Cup 1999 benchmark. 
    The dataset contains 43 features describing individual network connections, with labels indicating the traffic type (Normal, DoS, Probe, R2L).

    \noindent Our analysis is structured around four main tasks:
    \begin{enumerate}
        \item \textbf{Dataset characterization and preprocessing} (\Cref{sec:task1}): We explore the dataset structure, handle categorical and numerical features, and analyze per-class feature distributions through heatmaps and correlation analysis.
        \item \textbf{Shallow anomaly detection} (\Cref{sec:task2}): We evaluate One-Class SVM under three training configurations --- normal data only (novelty detection), all data (outlier detection), and mixed contamination levels --- quantifying the impact of the $\nu$ parameter and training data composition.
        \item \textbf{Deep anomaly detection} (\Cref{sec:task3}): We train an autoencoder on normal data and explore three detection strategies: reconstruction error thresholding, OC-SVM on bottleneck embeddings, and OC-SVM on PCA-reduced features.
        \item \textbf{Unsupervised clustering} (\Cref{sec:task4}): We apply K-Means and DBSCAN to assess whether purely unsupervised approaches can identify attack patterns, using silhouette analysis, t-SNE visualization, and cluster purity metrics.
    \end{enumerate}

    All experiments are conducted in Google Colab using scikit-learn and PyTorch, with an 80/20 stratified split performed before preprocessing to prevent data leakage.
