% task2.tex — Shallow Anomaly Detection with One-Class SVM

% Section Title
\section{SHALLOW ANOMALY DETECTION} \label{sec:task2}

    In this section, we investigate One-Class Support Vector Machine (OC-SVM) for anomaly detection under three training configurations: normal data only, all data, and mixed data with varying anomaly ratios.

    % --- OC-SVM Normal Only ---
    \subsection{OC-SVM with Normal Data Only (Novelty Detection)} \label{sec:ocsvm_normal}

        In novelty detection, the OC-SVM is trained exclusively on normal traffic, learning a tight boundary around the normal data manifold. 
        The key hyperparameter is $\nu$, which controls the upper bound on the fraction of training errors.

        \paragraph{$\nu$ parameter analysis.} Since the training data is purely normal, $\nu$ should reflect the expected fraction of noisy or edge-case normal samples. 
        We evaluated four values: $\nu \in \{0.001, 0.01, 0.05, 0.5\}$.

        \begin{table}[H]
            \centering
            \begin{tabular}{@{}lcc@{}}
                \toprule
                \textbf{$\nu$} & \textbf{Train F1-Macro} & \textbf{Test F1-Macro} \\
                \midrule
                0.001   & 0.9058 & 0.7382 \\
                0.01    & 0.9039 & 0.7401 \\
                \textbf{0.05} & \textbf{0.9026} & \textbf{0.7477} \\
                0.5 (default) & 0.6378 & 0.4647 \\
                \bottomrule
            \end{tabular}
            \caption{OC-SVM (Normal-Only): impact of $\nu$ on performance.}
            \label{tab:nu_comparison}
        \end{table}

        We selected $\nu = 0.05$ as the best estimate, as it achieves the highest test F1-Macro (0.7477). 
        This value allows the model to exclude approximately 5\% of normal samples as noise during training, creating a slightly looser boundary that generalizes better. 
        The default $\nu = 0.5$ is catastrophic for novelty detection, as it forces half of the normal training data to be treated as outliers, effectively collapsing the decision boundary.

    % --- OC-SVM All Data ---
    \subsection{OC-SVM with All Data (Outlier Detection)} \label{sec:ocsvm_all}

        When training on the full dataset (normal + anomalies), the OC-SVM operates as an \emph{outlier detector}. 
        We estimated $\nu$ as the true contamination rate: $\nu = 4305/15063 \approx 0.286$.

        \begin{table}[H]
            \centering
            \begin{tabular}{@{}lcc@{}}
                \toprule
                \textbf{Model} & \textbf{Test F1-Macro} & \textbf{Test Accuracy} \\
                \midrule
                Normal-Only ($\nu{=}0.05$) & \textbf{0.7477} & 0.7667 \\
                All-Data ($\nu{=}0.286$) & 0.6484 & 0.6656 \\
                \bottomrule
            \end{tabular}
            \caption{Comparison of Normal-Only vs.\ All-Data OC-SVM on the test set.}
            \label{tab:normal_vs_all}
        \end{table}

        The Normal-Only model outperforms the All-Data model by approximately 10 percentage points in F1-Macro. 
        This demonstrates that novelty detection (learning pure normality) is more effective than outlier detection (learning on contaminated data): the All-Data model's boundary is distorted by the inclusion of anomalies during training.

    % --- Sensitivity Analysis ---
    \subsection{Sensitivity Analysis: Impact of Anomaly Contamination} \label{sec:sensitivity}

        To quantify the effect of training data contamination, we trained OC-SVMs with increasing proportions of anomalous data: 0\%, 10\%, 20\%, 50\%, and 100\% of available anomalies. 
        For each configuration, $\nu$ was set proportionally to the anomaly ratio.

        \begin{table}[H]
            \centering
            \begin{tabular}{@{}rccc@{}}
                \toprule
                \textbf{Anomaly \%} & \textbf{$\nu$} & \textbf{Train F1-Macro} & \textbf{Test F1-Macro} \\
                \midrule
                0\% (Normal-Only)  & 0.01   & 0.9039 & \textbf{0.7401} \\
                10\%               & 0.038  & 0.6628 & 0.5137 \\
                20\%               & 0.074  & 0.6779 & 0.5589 \\
                50\%               & 0.167  & 0.7226 & 0.5975 \\
                100\% (All-Data)   & 0.286  & 0.7364 & 0.6482 \\
                \bottomrule
            \end{tabular}
            \caption{Sensitivity analysis: impact of anomaly contamination on OC-SVM performance.}
            \label{tab:sensitivity}
        \end{table}

        The results reveal a non-monotonic pattern. 
        Small contamination levels (10--20\%) cause the most damage: the model receives mixed signals about what constitutes ``normal'' behavior, but $\nu$ is too small to properly account for the anomalies. 
        As contamination increases to 50--100\%, performance partially recovers because $\nu$ grows large enough to capture the true anomaly proportion. 
        However, no contaminated model matches the purely normal training baseline.

    % --- Robustness ---
    \subsection{Robustness Evaluation} \label{sec:robustness}

        We compared three representative models on the test set to assess robustness:

        \begin{table}[H]
            \centering
            \begin{tabular}{@{}lcc@{}}
                \toprule
                \textbf{Model} & \textbf{Test F1-Macro} & \textbf{Most Missed Attack} \\
                \midrule
                Normal-Only ($\nu{=}0.05$) & \textbf{0.7477} & DoS (609 missed) \\
                All-Data ($\nu{=}0.286$)   & 0.6484 & DoS (1,074 missed) \\
                10\% Anomalies              & 0.5137 & DoS (1,989 missed) \\
                \bottomrule
            \end{tabular}
            \caption{Robustness evaluation on the test set: three OC-SVM configurations.}
            \label{tab:robustness}
        \end{table}

        \paragraph{Per-attack confusion analysis.} DoS attacks generate the most false negatives across all models, both in absolute count and due to the distribution shift (DoS increased from 15.5\% to 45.2\% in the test set). 
        R2L attacks are the hardest to detect proportionally, as their features closely mimic normal traffic patterns. 
        The Normal-Only model shows the best overall balance between precision (fewer false alarms) and recall (catching more attacks).
